---
title: "Bigleaf package use"
author: "Ronny Hernández Mora"
date: "2/28/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(bigleaf)
```

## Objetive

To understand the steps in order to calculate several values with help from the
functions from the bigleafR package.

# Data

Package contains an example data set
```{r}
data_example <- DE_Tha_Jun_2014
```

There is a need of ancillary data as:

```{r}
LAI <- 7.6 # leaf area index
zh <- 26.5 # average vegetation height (m)
zr <- 42 # sensor height (m)
Dl <- 0.01 # leaf characteristic dimension (m)
```

## Units

For each function in the package, we need to validate that in our dataset we
have the same units. In the documentation of each function there is information
about the variables needed as arguments in the functions and the units of
measurement.

## Variable names

Las funciones del paquete están para trabajar con los mismos nombres del 
conjunto de datos que viene de ejemplo, por lo que si queremos usar la funcion
sin dar los argumentos, nuestro conjunto  de datos debe de tener los mismos
nombres.

# Filtrado de datos

```{r}
tha_filtered <- filter.data(data, 
                             quality.control = TRUE,
                             vars.qc = c("LE", "H", "NEE", "Tair", "VPD", "wind"),
                             quality.ext = "_qc", 
                             good.quality = c(0, 1), 
                             missing.qc.as.bad = TRUE)
```

# Para calcular potential evapotranspiration

```{r}
potential.ET(tha_filtered, G = "G", approach = "Priestley-Taylor")
```











