---
title: "Evapotranspiration calculation"
subtitle: "EAS 520"
author: "Ronny Hernández Mora"
date: "2/27/2021"
always_allow_html: yes
output:
  rmdformats::downcute:
    code_folding: hide
    fig_width: 12
    fig_height: 8
    use_bookdown: true
editor_options:
  chunk_output_type: console
---

```{r setup-import-clean data, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(dplyr)
library(lubridate)
library(janitor)
library(readr)
library(tidyr)
library(stringr)
library(ggplot2)
library(skimr)
library(feather)
library(bigleaf)

# Import data

## Temperature
temperature_sensors <- read_csv("data/COSTA RICA - Santa Rosa National Park, Guanacaste 2013-06-06_2020-05-11_sensors.csv",
                    col_types = cols(`Air Temperature (34810)` = col_number(), 
                                     `Air Temperature (9861197)` = col_number(), 
                                     `Air Temperature (248)` = col_number(), 
                                     `Air Temperature (1)` = col_number(), 
                                     `Air Temperature (2)` = col_number())) %>% 
  clean_names() %>% 
  rename("temp_principe_wsn_pheno_34810" = "air_temperature_34810",
         "temp_perros_tower_hobo" = "air_temperature_9861197",
         "temp_phenology_tower_hobo" = "air_temperature_9861198",
         "temp_principe_wsn_pheno_248" = "air_temperature_248",
         "temp_carbon_tower_net_radiometer" = "air_temperature_1",
         "temp_principe_tower_campbell" = "air_temperature_2"
  ) %>% 
   pivot_longer(cols = c("temp_principe_wsn_pheno_34810",
                        "temp_perros_tower_hobo",
                        "temp_phenology_tower_hobo",
                        "temp_principe_wsn_pheno_248",
                        "temp_carbon_tower_net_radiometer",
                        "temp_principe_tower_campbell"
                        ),
               names_to = "dataset") %>% 
  filter(!is.na(value),
         value > 0) %>% 
  separate(time_stamp, into = c("date", "time"), sep = " ") %>% 
  mutate(time = str_sub(time, start = 1, end = 5)) %>% 
  unite(col = date_time, date, time, sep = " ") %>% 
  mutate(date_time = ymd_hm(date_time)) %>% 
  rename("temperature" = "value") %>% 
  # Llevar los tiempos a periodos de 30 minutos
  # Ver chunk con notas sobre pruebas realizadas
  mutate(date_time = ceiling_date(date_time, "30 mins"))


## Read carbon tower data with error removal
carbon_tower_eddy_error_removal <- read_csv("data/COSTARICA-SantaRosaNationalPark,Guanacaste_CarbonTower(New)_EddyCo._CampbellCR3000(1)_20130607_20190512_error_removal.csv") %>% 
  clean_names() %>% 
  select(date_time, sensible_heat_flux, co2_flux, latent_heat_flux,
         mean_co2_concentration, mean_h2o_vapour_concentration, 
         mean_moist_air_density, mean_thermocouple_temp) %>% 
  mutate(tower = "carbon_tower") %>% 
  # Set seconds to 0
  separate(date_time, into = c("date", "time"), sep = " ") %>% 
  mutate(time = str_sub(time, start = 1, end = 5)) %>% 
  unite(col = date_time, date, time, sep = " ") %>% 
  mutate(date_time = ymd_hm(date_time)) 

## Read principe tower data with error removal
principe_tower_eddy_error_removal <- read_csv("data/COSTARICA-SantaRosaNationalPark,Guanacaste_PrincipeTower(New)_EddyCo._CampbellCR3000(2)_20151213_20190622_error_removal.csv") %>% 
  clean_names() %>% 
  select(date_time, sensible_heat_flux, co2_flux, latent_heat_flux,
         mean_co2_concentration, mean_h2o_vapour_concentration, 
         mean_moist_air_density, mean_thermocouple_temp) %>% 
  mutate(tower = "principe_tower") %>% 
  # Set seconds to 0
  separate(date_time, into = c("date", "time"), sep = " ") %>% 
  mutate(time = str_sub(time, start = 1, end = 5)) %>% 
  unite(col = date_time, date, time, sep = " ") %>% 
  mutate(date_time = ymd_hm(date_time)) 


### Bind towers data sets
towers <- bind_rows(carbon_tower_eddy_error_removal,
                    principe_tower_eddy_error_removal
                    ) %>% 
  # Llevar los tiempos a periodos de 30 minutos
  # Ver chunk con notas sobre pruebas realizadas
  mutate(date_time = ceiling_date(date_time, "30 mins"))
```

# Consideraciones:

 - Para los datos de temperatura, si tenemos muchos valores por minuto, por
 hora por dia, agruparlos y sacar un promedio.
 - Luego, estos valores promediados por minutos, promediarlos por hora o por
 día para hacer el calculo.
 - Tenemos que unir los conjuntos de datos que contienen latent_heat_flux y
 temperatura para conocer cual es el periodo de tiempo mas largo posible que
 tenemos.
 - Con evapotranspiracion calculado, sacar conclusiones
 
# Date ranges data visualization

```{r date range visualization}
# Latent Heat Flux in time
towers %>% 
  mutate(date = ymd_hms(date_time)) %>%
  ggplot(aes(x = date_time, y = latent_heat_flux, color = tower)) +
  geom_jitter(alpha = 1, size = 0.7) +
  scale_y_continuous(breaks = seq(-4000, 3000, by = 500)) + #
  scale_x_datetime(date_labels = "%b%Y", breaks = "months") +
  theme_light(base_size = 16) +
  theme(axis.text.x = element_text(angle = 75, h = 1)) +
  labs(title = "Latent Heat Flux for all observations available from both towers",
       subtitle = "Without filtering outliers",
       x = "Date", 
       y = "Latent Heat Flux (W m-2)",
       color = "Tower")

# Temperature in time
temperature_sensors %>% 
  mutate(date = ymd_hms(date_time)) %>%
  ggplot(aes(x = date_time, y = temperature)) +
  geom_point(alpha = 0.5) +
  scale_y_continuous(breaks = seq(10, 40, by = 2)) +
  scale_x_datetime(date_labels = "%b%Y", breaks = "months") +
  theme_light(base_size = 16) +
  theme(axis.text.x = element_text(angle = 75, h = 1)) +
  labs(title = "Temperature (C)",
       subtitle = "All the datapoints available from sensor dataset",
       x = "Date", y = "Temperature")
```

# Evapotranspiration calculation

```{r evapotranspiration calculation}
# Unidades = kg H2O m−2 d−1

# Obtain mean temperaure value grouped by date_time
temp_mean <- temperature_sensors %>% 
  group_by(date_time) %>% 
  summarise(
    mean_sensors_temp = mean(temperature, na.rm = TRUE)
  )

# Test to check if we have unique date_time values
stopifnot(nrow(temp_mean) == length(unique(temp_mean$date_time)))

# Join temperature means with towers dataset
evapotranspiration <- towers %>% 
  # distinct(date_time, .keep_all = TRUE)
  # Filter outliers 
  filter(latent_heat_flux > -500,
         latent_heat_flux < 1000) %>% 
  # Este inner join me excluye valores repetidos de latent_heat_flux
  inner_join(temp_mean, by = "date_time") %>% 
  filter(!is.na(latent_heat_flux)) %>% 
  # Calculate evapotranspiration
  mutate(evapotranspiration_kg = LE.to.ET(latent_heat_flux,
                                          mean_sensors_temp )) %>% 
  # Transform units
  mutate(evapotranspiration_mol = kg.to.mol((evapotranspiration_kg) * 1000))

# Review towers repeated date_times
# check_towers <- towers %>% 
```

<!-- Pruebas con cambios de minutos para cerrarlos en media hora -->

```{r, eval = FALSE}
# date validation between datasets
# check_temp <- temp_mean %>% 
#   select(date_time, mean_sensors_temp) %>% 
#   mutate(dataset = "sensors",
#          fecha = date_time)
# 
# check_towers <- towers %>% 
#   select(date_time, latent_heat_flux) %>% 
#   mutate(dataset = "sensors",
#          fecha = date_time)
# 
# check_left <- left_join(check_temp, check_towers, by = "date_time")
# check_right <- right_join(check_temp, check_towers, by = "date_time")
# check_inner <- inner_join(check_temp, check_towers, by = "date_time")
# 
# 
# # Tengo los minutos de temperatura a 14 o 44 y no cerrados a la hora o media hora
# x <- ymd_hms("2009-08-03 12:01:59.23")
# floor_date(x, "minute")
# 
# floor_date(x, "hour")
# 
# x <- ymd_hms("2013-06-09 21:14:00")
# ceiling_date(x, "30 mins")
# 
# x <- ymd_hms("2013-06-09 21:00:00")
# ceiling_date(x, "30 mins")
# 
# # Prueba cambio de hora
# check_temp %>% 
#   slice(100:120) %>% 
#   mutate(date_time = ymd_hms(date_time)) %>% 
#   # separate(date_time, into = c("date", "time"), sep = " ") %>% 
#   # mutate(time = hms(time),
#   #        date = ymd(date)) %>% 
#   # mutate(minutos = ifelse(minute(time) > 30, 
#   #                         ceiling_date(time, "30 mins"),
#   #                         floor_date(time, "hour"))
#   # ) %>%
#   mutate(minutos = ifelse(minute(date_time) > 30, 
#                           ceiling_date(date_time, "30 mins"),
#                           floor_date(date_time, "hour"))
#   ) %>% 
#   mutate(fecha_check = as_date(minutos, origin = lubridate::origin)) %>% View()
#  # mutate(fecha_check = ymd_hms(minutos)) %>% View()
# 
# check_temp %>% 
#   slice(100:120) %>% 
#   mutate(minutos = ceiling_date(date_time, "30 mins")
#   ) %>% View()
```

```{r}
evapotranspiration %>% 
  mutate(date = ymd_hms(date_time)) %>%
  ggplot(aes(x = date_time, y = evapotranspiration_kg)) +
  geom_jitter(alpha = 0.06, size = 0.7) +
  scale_y_continuous() + #breaks = seq(10, 40, by = 2)
  scale_x_datetime(date_labels = "%b%Y", breaks = "months") +
  theme_light(base_size = 16) +
  theme(axis.text.x = element_text(angle = 75, h = 1)) +
  labs(title = "Evapotranspiration",
       subtitle = "All the datapoints from available",
       x = "Date", y = "Evapotranspiration (kg H2O m−2 d−1)")
```

## Data aggregation

In personal communication with Juergen Knauer (The developer of the Bigleaf R 
package) says that if we want to see an effect per season, it will be better if 
the data is aggregated per date.

```{r grouped evapotranspiration}
evapotranspiration_grouped <- evapotranspiration %>% 
  mutate(date = as.Date(date_time)) %>% 
  group_by(ymd(date)) %>% 
  summarise(
    mean = mean(evapotranspiration_mol, na.rm = TRUE),
    total = sum(evapotranspiration_mol, na.rm = TRUE)
  ) %>% 
  rename("date" = `ymd(date)`)
```


```{r, eval = FALSE}
# Plot sum of evapotranspiration
evapotranspiration_grouped %>% 
  mutate(date = ymd(date)) %>%
  ggplot(aes(x = date, y = total)) +
  geom_point(alpha = 0.5) +
  scale_x_date(date_labels = "%b-%d-%Y", breaks = "months") +
  theme_light(base_size = 21) +
  theme(axis.text.x = element_text(angle = 75, h = 1)) +
  labs(title = "Sum of evapotranspiration per date",
       subtitle = "Al the datapoints available",
       x = "Date", y = "Evapotranspiration mmol H2O m-2 s-1")

# Plot mean of evapotranspiration
evapotranspiration_grouped %>% 
  mutate(date = ymd(date)) %>%
  ggplot(aes(x = date, y = mean)) +
  geom_point(alpha = 0.5) +
  geom_smooth() +
  scale_x_date(date_labels = "%b-%d-%Y", breaks = "2 months") +
  theme_light(base_size = 21) +
  theme(axis.text.x = element_text(angle = 75, h = 1)) +
  labs(title = "Mean of evapotranspiration per date",
       subtitle = "Al the datapoints available",
       x = "Date", y = "Evapotranspiration mmol H2O m-2 s-1")
```





### References

https://stackoverflow.com/questions/9968975/make-the-background-of-a-graph-different-colours-in-different-regions
https://stackoverflow.com/questions/50339909/shade-background-of-a-ggplot-chart-using-geom-rect-with-categorical-variables#50340075





